---
title: "Blood Pressure and Cardiometabolic Risk in U.S. Adults"
author: "Andrew Qian"
date: last-modified
format: 
  dashboard:
    theme: pulse
    orientation: rows
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
library(tidyverse)

demo_raw <- readRDS("data/P_DEMO.Rds")
bpx_raw <- readRDS("data/P_BPXO.Rds")
bmx_raw <- readRDS("data/P_BMX.Rds")
diq_raw <- readRDS("data/P_DIQ.Rds")
smq_raw <- readRDS("data/P_SMQ.Rds")

demo_sel <- demo_raw |> select(SEQN, RIDSTATR, RIDAGEYR, RIAGENDR, RIDRETH3, DMDEDUC2)
bpx_sel <- bpx_raw |> select(SEQN, BPXOSY1, BPXOSY2, BPXOSY3)
bmx_sel <- bmx_raw |> select(SEQN, BMXBMI)
diq_sel <- diq_raw |> select(SEQN, DIQ010)
smq_sel <- smq_raw |> select(SEQN, SMQ020)

dash_data <- demo_sel |> 
  left_join(bpx_sel, by = "SEQN") |> 
  left_join(bmx_sel, by = "SEQN") |> 
  left_join(diq_sel, by = "SEQN") |> 
  left_join(smq_sel, by = "SEQN") |> 
  filter(RIDSTATR == "Both interviewed and MEC examined", RIDAGEYR >= 21, RIDAGEYR < 80) |> 
  mutate(
    sbp = rowMeans(across(c(BPXOSY1, BPXOSY2, BPXOSY3)), na.rm = TRUE),
    sex = factor(RIAGENDR),
    race_eth = factor(RIDRETH3),
    education = factor(DMDEDUC2),
    smoker = factor(SMQ020),
    diabetes = factor(DIQ010)
  ) |> 
  mutate(
    education = na_if(education, "Refused"), education = na_if(education, "Don't Know"),
    education = fct_collapse(education,
      "Less than HS" = c("Less than 9th grade", "9-11th grade (Includes 12th grade with no diploma)"),
      "HS Grad" = "High school graduate/GED or equivalent",
      "Some College" = "Some college or AA degree",
      "College Grad" = "College graduate or above"),
    education = droplevels(education),
    smoker = na_if(smoker, "Refused"), smoker = na_if(smoker, "Don't know"), smoker = droplevels(smoker),
    diabetes = na_if(diabetes, "Don't know"), diabetes = na_if(diabetes, "Borderline"), diabetes = droplevels(diabetes)
  ) |> 
  drop_na(sbp, BMXBMI, sex, race_eth, education, smoker, diabetes)

set.seed(431)
train_dash <- dash_data |> slice_sample(prop = 0.75) |> mutate(log_sbp = log(sbp))
test_dash <- anti_join(dash_data, train_dash, by = "SEQN") |> mutate(log_sbp = log(sbp))

small_model_dash <- lm(log_sbp ~ BMXBMI + RIDAGEYR + sex + race_eth + education, data = train_dash)
test_dash <- test_dash |> mutate(pred_small = exp(predict(small_model_dash, newdata = test_dash)))
```

# Study 1: Group Comparisons

## Row {height="15%"}

::: {.card}
This section compares systolic blood pressure and cardiometabolic risk factors across demographic and lifestyle groups in a sample of 6,387 U.S. adults (NHANES 2017–March 2020).
:::

## Row

```{r}
#| title: "SBP by Sex"
ggplot(dash_data, aes(x = sex, y = sbp)) +
  geom_boxplot(fill = "cornflowerblue") +
  labs(x = "Sex", y = "Systolic BP (mm Hg)",
       caption = "Men show higher median and mean SBP than women (126.5 vs. 121.1 mmHg, p < 0.001).")
```

```{r}
#| title: "SBP by Race/Ethnicity"
ggplot(dash_data, aes(x = race_eth, y = sbp)) +
  geom_boxplot(fill = "cornflowerblue") +
  labs(x = "", y = "Systolic BP (mm Hg)",
       caption = "Non-Hispanic Black adults show notably higher SBP than every other group (p < 0.001).") +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

## Row

```{r}
#| title: "Smoking and Diabetes Diagnosis"
ggplot(dash_data, aes(x = smoker, fill = diabetes)) +
  geom_bar(position = "fill") +
  labs(x = "Ever Smoked 100+ Cigarettes", y = "Proportion", fill = "Diabetes",
       caption = "Smokers show a higher diabetes rate than non-smokers (18.2% vs. 13.5%, p < 0.001).")
```

::: {.card}
### Key Findings

- Men had significantly higher mean systolic blood pressure than women (p < 0.001).
- Non-Hispanic Black adults had significantly higher mean systolic blood pressure than every other racial/ethnic group (p < 0.001).
- Smokers showed a notably higher diabetes rate than non-smokers (p < 0.001).
- Education level and race/ethnicity were significantly associated (χ² = 1046.4, p < 0.001).
:::

# Study 2: Predicting Blood Pressure

## Row {height="15%"}

::: {.card}
This section builds a regression model to test how well BMI predicts systolic blood pressure, adjusting for age, sex, race/ethnicity, and education, in a sample of 7,029 U.S. adults.
:::

## Row

```{r}
#| title: "Distribution of log(Systolic Blood Pressure)"
ggplot(train_dash, aes(x = log_sbp)) +
  geom_histogram(binwidth = 0.02, fill = "cornflowerblue", color = "black") +
  labs(x = "log(Systolic Blood Pressure)", y = "Frequency",
       caption = "Log transformation was used to correct for right-skew in raw SBP.")
```

```{r}
#| title: "Predicted vs. Observed SBP (Test Sample)"
ggplot(test_dash, aes(x = pred_small, y = sbp)) +
  geom_point(alpha = 0.4) +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  labs(x = "Predicted SBP (mm Hg)", y = "Observed SBP (mm Hg)",
       caption = "Points near the dashed line indicate accurate predictions on unseen data.")
```

## Row

::: {.card}
### Research Question

How effectively does BMI predict systolic blood pressure in U.S. adults, after adjusting for age, sex, race/ethnicity, education, smoking status, and diabetes diagnosis?

### Key Findings

- After adjusting for other factors, **BMI was only marginally associated** with blood pressure (p = 0.082) — its raw association appears largely explained by age, sex, race, and education.
- **Age was the strongest predictor** of systolic blood pressure by far.
- A simplified model performed **just as well** as the full model (R² ≈ 0.225 for both), on both training and test data.
:::